High-Dimensional Regularized Discriminant Analysis

### 2016-01-07 reports/paper-hdrda.Rnw
Some edits from Phil.
DMY is going to have a look at this version now.

\usepackage{graphicx, amsmath, amssymb, bm, mathtools, amsthm}
\usepackage{natbib, setspace}
\usepackage{algorithm2e}

@@

\citeauthor{Friedman:1989tm} proposed the popular regularized discriminant
analysis (RDA) classifier that utilizes a biased covariance-matrix estimator to
partially pool the sample covariance matrices from linear and quadratic
disriminant analysis before shrinking the resulting estimator towards a scaled
identity matrix. The RDA classifier's two tuning parameters are typically
estimated via a computationally burdensome cross-validation procedure from a
grid of candidate values. We reformulate the RDA classifier in the small-sample,
high-dimensional setting and then show that the classification decision rule is
equivalent to one in a subspace having a much lower dimension. As a result, the
dimension reduction involved yields a substantial reduction in computation
during model selection. Our parameterization also offers interpretability
previously lacking with the RDA classifier. We demonstrate with three artificial
and four real high-dimensional data sets that our proposed classifier is often
superior to several recently proposed sparse and regularized classifiers in
terms of classification accuracy. Finally, we provide an implementation of our
proposed classifier in the {\tt sparsediscrim} R package, available on CRAN.

@@

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers}
employed in a variety of disciplines, such as medical research, facial
recognition, remote sensing, and chemical analysis
\citep{Ching:2012fu,Li:2012ev,Zhang:2010va}. Despite the advances in
classification to facilitate technological advancements, well-established
classifiers often degrade in accuracy with modern, high-dimensional data sets
when the number of features $p$ exceeds the total sample size $N$, due to the
\emph{curse of dimensionality} \citep{Bellman:1961tn}. For instance, linear
discriminant analysis (LDA) and quadratic discriminant analysis (QDA) are
popular because of their simplicity, speed of computation, and
interpretability. However, both methods are incalculable when $p > N$ because
the class sample covariance matrices are singular \citep*{Bouveyron:2007gx}.

As a result, researchers have emphasized three standard approaches to modifying
the \emph{LDA} and \emph{QDA} classifiers for application to high-dimensional
data: 1) covariance-matrix regularization, 2) dimension reduction, and 3)
parameter restrictions. Covariance-matrix regularization methods are often based
on the well-known ridge-regression approach of \cite{Hoerl:1970cd}.
\cite{Ching:2012fu}, \cite{Zhang:2010va}, \cite{Ye:2009gd}, \cite{Ji:2008wp},
\cite*{Guo:2007te}, \cite*{Srivastava:2007ud}, \cite{Ye:2006vx}, and
\cite{Ye:2005uu} have proposed regularization methods to improve the estimation
of the class covariance matrices by adjusting the eigenvalues of the sample
covariance matrices to ensure positive definiteness. Alternatively, dimension
reduction typically transforms the original data in the high-dimensional space
to a subspace having a much smaller dimension such that redundant and noisy
information is removed
\citep{Fan:2012iq,Ye:2009gd,Bouveyron:2007gx,Ye:2006vx,Ye:2005hd,Antoniadis:2003bk}. Finally,
\cite{Mai:2012bf}, \cite{Fan:2012iq}, \cite{Merchante:2012vk},
\cite{Clemmensen:2011kr}, \cite{Huang:2010ju}, and \cite{Dudoit:2002ev} have
proposed restrictions on the population covariance matrices to reduce the number
of parameters that must be estimated. For instance, \cite{Dudoit:2002ev} have
assumed the features within each class are independent so that the population
covariance matrices are diagonal, resulting in positive-definite sample
covariance matrices. For excellent overviews of the regularization problem in
discriminant analysis and other techniques, see \cite{Murphy:2012uq} and
\cite{Mkhadri:1997gy}.

Here, we present a new classification method intended for the $p > N$ case that
builds on the classic regularized discriminant analysis (\emph{RDA}) proposed by
\cite{Friedman:1989tm}. Our proposed high-dimensional \emph{RDA} (\emph{HDRDA})
classifier employs a biased covariance matrix estimator that partially pools the
individual sample covariance matrices from the \emph{QDA} classifier with the
pooled sample covariance matrix from the \emph{LDA} classifier. We then shrink
the resulting covariance-matrix estimator towards a scaled identity matrix to
ensure positive definiteness. By reparameterizing the \emph{RDA} classifier
similar to \cite{Hastie:2008dt} and \cite{Halbe:2007ea}, we first show the
pooling parameter in the \emph{HDRDA} classifier determines the contribution of
each training observation to the estimation of each class covariance matrix,
enabling interpretability that has been previously lacking with the \emph{RDA}
classifier \citep{Bensmail:1996fw}.

Next, we achieve a substantial reduction in dimension to reduce greatly the
computation during model selection by applying a reasoning similar to
\cite{Ye:2006tq}. We establish that the matrix operations corresponding to the
null space of the pooled sample covariance matrix are redundant and can be
discarded from the \emph{HDRDA} decision rule without loss of classificatory
information. Moreover, because the \emph{HDRDA} decision rule is invariant to
adjustments to the zero eigenvalues, the decision rule in the original feature
space is equivalent to a decision rule in a much lower dimension, such that the
matrix inverses and determinants of relatively small matrices can be rapidly
computed. Furthermore, we show that a large number of shrinkage methods that are
special cases of the \emph{HDRDA} classifier have no effect on the zero
eigenvalues of the covariance-matrix estimators. Such methods include work from
\cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier when $p \gg N$. We then study the classification
performance of the \emph{HDRDA} classifier on four real high-dimensional data
sets along with a simulation experiment that generalizes that initially
conducted by \cite{Guo:2007te}. We demonstrate that the \emph{HDRDA} classifier
often attains superior classification accuracy to several recent classifiers
designed for small-sample, high-dimensional data from \cite{Tong:2012hw},
\cite{Witten:2011kc}, \cite{Pang:2009ik}, and \cite{Guo:2007te}. We also include
as a benchmark the random forest from \cite{Breiman:2001fb} because
\cite{FernandezDelgado:2014ul} has concluded that the random forest is often
superior to other classifiers in benchmark studies.

@@

To facilitate our discussion of covariance-matrix regularization and regularized
discriminant analysis, we require the following notation. Let $\mathbb{R}_{a
  \times b}$ denote the matrix space of all $a \times b$ matrices over the real
field $\mathbb{R}$. Denote by $\bm I_m$ the $m \times m$ identity matrix, and
let $\bm 0_{m \times p}$ be the $m \times p$ matrix of zeros, such that $\bm
0_m$ is understood to denote $\bm 0_{m \times m}$. Define $\bm 1_m \in
\mathbb{R}_{m \times 1}$ as a vector of ones. Let $\bm A \oplus \bm B$ denote
the direct sum of $\bm A \in \mathbb{R}_{r \times r}$ and $\bm B \in
\mathbb{R}_{s \times s}$ \citep[Chapter 1]{Lutkepohl:1996uz}. Let $\bm A^{T}$,
$\bm A^+$, and $\mathcal{N}(\bm A) = \{\bm z \in \mathbb{R}_{p \times 1} : \bm A
\bm z = \bm 0_{m \times 1}\}$ denote the transpose, the Moore-Penrose
pseudoinverse, and the null space of $\bm A \in \mathbb{R}_{m \times p}$,
respectively. Denote by $\mathbb{R}_{p \times p}^{>}$ the cone of real $p \times
p$ positive-definite matrices. Similarly, let $\mathbb{R}_{p \times p}^{\ge}$
denote the cone of real $p \times p$ positive-semidefinite matrices. Let
$V^{\perp} = \{\bm y \in \mathbb{R}_{p \times 1} : \bm y^{T} \bm v = 0$ for all
$\bm v \in V \}$ denote the orthogonal complement of a vector space $V \subset
\mathbb{R}_{p \times 1}$. For $c \in \mathbb{R}$, let $c^+ = 1/c$ if $c \ne 0$
and $0$ otherwise.

@@

In discriminant analysis we wish to assign an unlabeled vector $\bm x \in
\mathbb{R}_{p \times 1}$ to one of $K$ unique, known classes by constructing a
classifier from $N$ training observations. Let $\bm x_i = (x_{i1}, \ldots,
x_{ip}) \in \mathbb{R}_{p \times 1}$ be the $i$th observation $(i = 1, \ldots,
N)$ with true, unique membership $y_i \in \{\omega_1, \ldots, \omega_K\}$. We
assume that $(\bm x_i, y_i)$ is a realization from a mixture distribution
$p(\bm x) = \sum_{k=1}^K p(\bm x | \omega_k) p(\omega_k)$, where $p(\bm x |
\omega_k)$ is the probability density function (PDF) of the $k$th class and
$p(\omega_k)$ is the prior probability of class membership of the $k$th
class. We further assume that each of $p(\omega_k)$ is equal, $k = 1, \ldots,
K$.

Quadratic discriminant analysis is the optimal Bayesian decision rule with
respect to a $0-1$ loss function when $p(\bm x | \omega_k)$ is the probability
density function of the multivariate normal distribution with known mean vectors
$\bm \mu_k \in \mathbb{R}_{p \times 1}$ and known covariance matrices $\bm
\Sigma_k \in \mathbb{R}_{p \times p}^{>}$, $k = 1, 2, \ldots, K$. Because $\bm
\mu_k$ and $\bm \Sigma_k$ are typically unknown, we assign an unlabeled
observation $\bm x$ to class $\omega_k$ with the sample quadratic discriminant
analysis classifier
\begin{align}
	D_{QDA}(\bm x) = \argmin_{k} (\bm x - \xbar_k)^{T}\widehat{\bm
    \Sigma}_k^{-1}(\bm x - \xbar_k) + \log |\widehat{\bm \Sigma}_k|, \label{eq:qda}
\end{align}
where $\xbar_k$ and $\widehat{\bm \Sigma}_k$ are the maximum likelihood
estimators (MLEs) of $\bm \mu_k$ and $\bm \Sigma_k$, respectively. If we assume
further that $\bm \Sigma_k = \bm \Sigma$, $k = 1, \ldots, K$, then the pooled
sample covariance matrix $\widehat{\bm \Sigma}$ is substituted for $\widehat{
  \bm \Sigma}_k$ in \eqref{eq:qda}, where
\begin{align}
	\widehat{\bm \Sigma} = N^{-1} \sum_{k=1}^K n_k \widehat{\bm \Sigma}_k \label{eq:pooled-cov}
\end{align}
is the MLE for $\bm \Sigma$. In this case \eqref{eq:qda} reduces to the sample
linear discriminant analysis classifier, and the log-determinant is omitted
because it is constant across the $K$ classes.

@@

Several regularization methods, such as the methods considered by
\cite{Xu:2009fl}, \citet*{Guo:2007te}, and \cite{Mkhadri:1995jp} have been
proposed in the literature to adjust the eigenvalues of $\widehat{\bm \Sigma}_k$
so that \eqref{eq:qda} is calculable and has reduced variability in the standard
bias-variance tradeoff. A common form of the covariance-matrix regularization
applies a shrinkage factor $\gamma > 0$, so that
\begin{align}
	\widehat{\bm \Sigma}_k(\gamma) = \widehat{\bm \Sigma}_k + \gamma \bm I_p, \label{eq:ridge-estimator}
\end{align}
similar to a method employed in ridge regression \citep{Hoerl:1970cd}. Equation
\eqref{eq:ridge-estimator} effectively \emph{shrinks} the sample covariance
matrix $\widehat{\bm \Sigma}_k$ toward $\bm I_p$, thereby increasing the
eigenvalues of $\widehat{\bm \Sigma}_k$ by $\gamma$. Specifically, the zero
eigenvalues are replaced with $\gamma$, so that \eqref{eq:ridge-estimator} is
positive definite. For additional covariance-matrix regularization methods, see
\cite{Ramey:2013ji}, \cite{Xu:2009fl}, and \cite{Ye:2009gd}.

@@

Here, we define the \emph{HDRDA} classifier by first formulating the
covariance-matrix estimator $\widehat{\bm \Sigma}_k(\lambda)$ and demonstrating
its clear interpretation as a linear combination of the crossproducts of the
training observations centered by their respective class sample means. We define
the convex combination
\begin{align}
  \widehat{\bm \Sigma}_k(\lambda) = (1 - \lambda) \widehat{\bm \Sigma}_k + \lambda
  \widehat{\bm \Sigma}, \quad k = 1, \ldots, K,\label{eq:sig-lambda-alternative}
\end{align}
where $\lambda \in [0, 1]$ is the \emph{pooling} parameter. By rewriting
\eqref{eq:sig-lambda-alternative} in terms of the observations $ \bm x_i$ $(i =
1, \ldots, N)$, each centered by its class sample mean, we attain a clear
interpretation of $\widehat{\bm \Sigma}_k(\lambda)$. That is,
\begin{align}
	\widehat{\bm \Sigma}_k(\lambda)
  &= \left( 1 - \lambda + \frac{\lambda n_k}{N} \right) \widehat{\bm \Sigma}_k +  \frac{\lambda}{N} \sum_{\substack{k' = 1\\k' \ne k}}^K n_{k'} \widehat{\bm \Sigma}_{k'} \nonumber \\
	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right)\sum_{i=1}^N
  I(y_i = k) \bm x_i \bm x_i^{T} +  \frac{\lambda}{N} \sum_{i=1}^N I(y_i \ne k)
  \bm x_i \bm x_i^{T} \nonumber \\
	&= \sum_{i=1}^N c_{ik}(\lambda) \bm x_i \bm x_i^{T},\label{eq:sig-lambda-alternative2}
\end{align}
where $c_{ik}(\lambda) = \lambda N^{-1} + (1 - \lambda)n_k^{-1}I(y_i = k)$.
From \eqref{eq:sig-lambda-alternative2}, $\lambda$ weights the contribution of
each of the $N$ observations in estimating $\bm \Sigma_k$ from all $K$ classes
rather than $n_k$ observations from a single class. As a result, we can
interpret \eqref{eq:sig-lambda-alternative2} as a covariance-matrix estimator
that borrows from $\widehat{\bm \Sigma}$ in \eqref{eq:pooled-cov} to estimate $
\bm \Sigma_k$.

@@

As we have discussed above, several eigenvalue adjustment methods have been
proposed to increase eigenvalues (approximately) equal to 0. To further improve
the estimation of $\bm \Sigma_k$ and to stabilize the estimator's inverse, we
define the eigenvalue adjustment of \eqref{eq:sig-lambda-alternative} as
\begin{align}
	\tilde{\bm \Sigma}_k = \alpha_k \widehat{\bm \Sigma}_k(\lambda) + \gamma \bm I_p,\label{eq:hdrda-cov}
\end{align}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. Thus, the \emph{pooling} parameter $\lambda$ controls the amount
borrowed from $\widehat{\bm \Sigma}$ to estimate $\bm \Sigma_k$, and the
\emph{shrinkage} parameter $\gamma$ determines the degree of eigenvalue
shrinkage. The choice of $\alpha_k$ allows for a flexible formulation of
covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1, \ldots,
K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}. Similarly,
if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form similar to the
\emph{RDA} classifier from \cite{Friedman:1989tm}. Substituting
\eqref{eq:hdrda-cov} into \eqref{eq:qda}, we define the \emph{HDRDA} classifier
\begin{align}
	D_{HDRDA}(\bm x) = \argmin_{k}  (\bm x - \xbar_k)^{T}\tilde{\bm \Sigma}_k^{+}(\bm x - \xbar_k)  + \log |\tilde{\bm \Sigma}_k|. \label{eq:hdrda}
\end{align}
For $\gamma > 0$, $\tilde{\bm \Sigma}_k$ is nonsingular such that $\tilde{\bm
  \Sigma}_k^{-1}$ can be substituted for $\tilde{\bm \Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\bm \Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\bm
  \Sigma}_k$. Following \cite{Friedman:1989tm}, we select $\lambda$ and $\gamma$
from a grid of candidate models via cross-validation. We provide an
implementation of \eqref{eq:hdrda} in the {\tt hdrda} function in the {\tt
  sparsediscrim} R package, available on CRAN.

@@

Next, we establish properties of the covariance-matrix estimator and the
decision rule employed in the \emph{HDRDA} classifier. By doing so, we
demonstrate that \eqref{eq:hdrda} lends itself to a more efficient
calculation. We decompose \eqref{eq:hdrda} into a sum of two components, where
the first summand consists of matrix operations applied to low-dimensional
matrices and the second summand corresponds to the null space of $\widehat{\bm
  \Sigma}$ in \eqref{eq:pooled-cov}. We show that the matrix operations
performed on the null space of $\widehat{\bm \Sigma}$ yield constant quadratic
forms across all classes and can be omitted. For $p \gg N$, the constant
component involves determinants and inverses of high-dimensional matrices, and
by ignoring these calculations, a substantial reduction in computational costs
is achieved. Furthermore, a byproduct is that adjustments to the associated
eigenvalues have no effect on \eqref{eq:hdrda}. Lastly, we utilize the singular
value decomposition to calculate efficiently the eigenvalue decomposition of
$\widehat{\bm \Sigma}$, further reducing the computational costs of the
\emph{HDRDA} classifier.

@@

\subsection{Classification Results}

Similar to \cite{Witten:2011kc}, we report the average test error rates obtained
over 250 random training-test partitions in Table \ref{tab:microarray-results}.
Standard deviations of the test error rates are given in the parentheses. The
\emph{HDRDA} and Guo classifiers were superior in classification performance for
the majority of the simulations with the popular Guo classifier being the best
for the Burczysnki data set. Both \emph{HDRDA} methods yielded comparable
results.

>>>>

\usepackage{graphicx, amsmath, amssymb, bm, mathtools, amsthm, natbib, setspace, algorithm2e}

@@

\citeauthor{Friedman:1989tm} proposed the popular regularized discriminant
analysis (RDA) classifier that utilizes a biased covariance-matrix estimator to
partially pool the sample covariance matrices from linear and quadratic
disriminant analysis before shrinking the resulting estimator towards a scaled
identity matrix. The RDA classifier's two tuning parameters are typically
estimated via a computationally burdensome cross-validation procedure from a
grid of candidate values. We reformulate the RDA classifier in the small-sample,
high-dimensional setting and then show that the classification decision rule is
equivalent to a classifier in a subspace having a much lower dimension. As a
result, the dimension reduction involved yields a substantial reduction in
computation during model selection. Our parameterization also offers
interpretability previously lacking with the RDA classifier. We demonstrate with
three artificial and four real high-dimensional data sets that our proposed
classifier is often superior to several recently proposed sparse and regularized
classifiers in terms of classification accuracy. Finally, we provide an
implementation of our proposed classifier in the {\tt sparsediscrim} R package,
available on CRAN.

@@

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers}
employed in a variety of disciplines, such as medical research, facial
recognition, remote sensing, and chemical analysis
\citep*{Ching:2012fu,Li:2012ev,Zhang:2010va}. Despite the advances in
classification to facilitate technological advancements, well-established
classifiers often degrade in accuracy with modern, high-dimensional data sets
when the number of features $p$ exceeds the total sample size $N$, due to the
\emph{curse of dimensionality} \citep{Bellman:1961tn}. For instance, linear
discriminant analysis (\emph{LDA}) and quadratic discriminant analysis
(\emph{QDA}) are popular because of their simplicity, speed of computation, and
interpretability. However, both methods are incalculable when $p > N$ because
the class sample covariance matrices are singular \citep*{Bouveyron:2007gx}.

As a result, researchers have emphasized three standard approaches to modifying
the \emph{LDA} and \emph{QDA} classifiers for application to high-dimensional
data: 1) covariance-matrix regularization, 2) dimension reduction, and 3)
parameter restrictions. Covariance-matrix regularization methods are often based
on the well-known ridge-regression approach of \cite{Hoerl:1970cd}.
\cite{Ching:2012fu}, \cite{Zhang:2010va}, \cite{Ye:2009gd}, \cite{Ji:2008wp},
\cite*{Guo:2007te}, \cite*{Srivastava:2007ud}, \cite{Ye:2006vx}, and
\cite{Ye:2005uu} have proposed regularization methods to improve the estimation
of the class covariance matrices by adjusting the eigenvalues of the sample
covariance matrices to ensure positive definiteness. Alternatively, dimension
reduction typically transforms the original data in the high-dimensional space
to a subspace having a much smaller dimension such that redundant and noisy
information is removed
\citep*{Fan:2012iq,Ye:2009gd,Bouveyron:2007gx,Ye:2006vx,Ye:2005hd,Antoniadis:2003bk}.
Finally, \cite*{Mai:2012bf}, \cite{Fan:2012iq}, \cite*{Merchante:2012vk},
\cite*{Clemmensen:2011kr}, \cite*{Huang:2010ju}, and \cite*{Dudoit:2002ev} have
proposed restrictions on the population covariance matrices to reduce the number
of parameters that must be estimated. For instance, \cite{Dudoit:2002ev} have
assumed the features within each class are independent so that the population
covariance matrices are diagonal, resulting in positive-definite sample
covariance matrices. For excellent overviews of the regularization problem in
discriminant analysis and other techniques, see \cite{Murphy:2012uq} and
\cite*{Mkhadri:1997gy}.

Here, we present a new classification method intended for the $p > N$ case that
builds on the classic regularized discriminant analysis (\emph{RDA}) proposed by
\cite{Friedman:1989tm}. Our proposed high-dimensional \emph{RDA} (\emph{HDRDA})
classifier employs a biased covariance matrix estimator that partially pools the
individual sample covariance matrices from the \emph{QDA} classifier with the
pooled sample covariance matrix from the \emph{LDA} classifier. We then shrink
the resulting covariance-matrix estimator towards a scaled identity matrix to
ensure positive definiteness. By reparameterizing the \emph{RDA} classifier
similar to \cite*{Hastie:2008dt} and \cite{Halbe:2007ea}, we first show that the
pooling parameter in the \emph{HDRDA} classifier determines the contribution of
each training observation to the estimation of each class covariance matrix,
enabling interpretability that has been previously lacking with the \emph{RDA}
classifier \citep{Bensmail:1996fw}.

Next, we achieve a substantial reduction in dimension to reduce greatly the
computation during model selection by applying a reasoning similar to
\cite{Ye:2006tq}. We establish that the matrix operations corresponding to the
null space of the pooled sample covariance matrix are redundant and can be
discarded from the \emph{HDRDA} decision rule without loss of classificatory
information. Moreover, because the \emph{HDRDA} decision rule is invariant to
adjustments to the zero eigenvalues, the decision rule in the original feature
space is equivalent to a decision rule in a lower dimension, such that matrix
inverses and determinants of relatively small matrices can be rapidly computed.
Furthermore, we show that several shrinkage methods that are special cases of
the \emph{HDRDA} classifier have no effect on the zero eigenvalues of the
covariance-matrix estimators. Such methods include work from
\cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

We provide an efficient algorithm that expedites the model-selection process for
the \emph{HDRDA} classifier when $p \gg N$. We then study the classification
performance of the \emph{HDRDA} classifier on four real high-dimensional data
sets along with a simulation design that generalizes the experiments initially
conducted by \cite{Guo:2007te}. We demonstrate that the \emph{HDRDA} classifier
often attains superior classification accuracy to several recent classifiers
designed for small-sample, high-dimensional data from \cite*{Tong:2012hw},
\cite{Witten:2011kc}, \cite*{Pang:2009ik}, and \cite{Guo:2007te}. We also
include as a benchmark the random forest from \cite{Breiman:2001fb} because
\cite*{FernandezDelgado:2014ul} has concluded that the random forest is often
superior to other classifiers in benchmark studies.

@@

To facilitate our discussion of covariance-matrix regularization and regularized
discriminant analysis, we require the following notation. Let $\mathbb{R}_{a
  \times b}$ denote the matrix space of all $a \times b$ matrices over the real
field $\mathbb{R}$. Denote by $\bm I_m$ the $m \times m$ identity matrix, and
let $\bm 0_{m \times p}$ be the $m \times p$ matrix of zeros, such that $\bm
0_m$ is understood to denote $\bm 0_{m \times m}$. Define $\bm 1_m \in
\mathbb{R}_{m \times 1}$ as a vector of ones. Let $\bm A \oplus \bm B$ denote
the direct sum of $\bm A \in \mathbb{R}_{r \times r}$ and $\bm B \in
\mathbb{R}_{s \times s}$ \citep[Chapter 1]{Lutkepohl:1996uz}. Let $\bm A^{T}$,
$\bm A^+$, and $\mathcal{N}(\bm A)$ denote the transpose, the Moore-Penrose
pseudoinverse, and the null space of $\bm A \in \mathbb{R}_{m \times p}$,
respectively. Denote by $\mathbb{R}_{p \times p}^{>}$ the cone of real $p \times
p$ positive-definite matrices. Similarly, let $\mathbb{R}_{p \times p}^{\ge}$
denote the cone of real $p \times p$ positive-semidefinite matrices. Let
$V^{\perp}$ denote the orthogonal complement of a vector space $V \subset
\mathbb{R}_{p \times 1}$. For $c \in \mathbb{R}$, let $c^+ = 1/c$ if $c \ne 0$
and $0$ otherwise.

@@

In discriminant analysis we wish to assign an unlabeled vector $\bm x \in
\mathbb{R}_{p \times 1}$ to one of $K$ unique, known classes by constructing a
classifier from $N$ training observations. Let $\bm x_i = (x_{i1}, \ldots,
x_{ip}) \in \mathbb{R}_{p \times 1}$ be the $i$th observation $(i = 1, \ldots,
N)$ with true, unique membership $y_i \in \{\omega_1, \ldots, \omega_K\}$. We
assume that $(\bm x_i, y_i)$ is a realization from a mixture distribution $p(\bm
x) = \sum_{k=1}^K p(\bm x | \omega_k) p(\omega_k)$, where $p(\bm x | \omega_k)$
is the probability density function (PDF) of the $k$th class and $p(\omega_k)$
is the prior probability of class membership of the $k$th class. We further
assume $p(\omega_k) = p(\omega_l)$, $1 \le k, l \le K$, $k \ne l$.

The \emph{QDA} classifier is the optimal Bayesian decision rule with respect to
a $0-1$ loss function when $p(\bm x | \omega_k)$ is the PDF of the multivariate
normal distribution with known mean vectors $\bm \mu_k \in \mathbb{R}_{p \times 1}$
and known covariance matrices $\bm \Sigma_k \in \mathbb{R}_{p \times p}^{>}$,
$k = 1, 2, \ldots, K$. Because $\bm \mu_k$ and $\bm \Sigma_k$ are typically
unknown, we assign an unlabeled observation $\bm x$ to class $\omega_k$ with the
sample \emph{QDA} classifier
\begin{align}
	D_{QDA}(\bm x) = \argmin_{k} (\bm x - \xbar_k)^{T}\widehat{\bm
    \Sigma}_k^{-1}(\bm x - \xbar_k) + \log |\widehat{\bm \Sigma}_k|, \label{eq:qda}
\end{align}
where $\xbar_k$ and $\widehat{\bm \Sigma}_k$ are the maximum likelihood
estimators (MLEs) of $\bm \mu_k$ and $\bm \Sigma_k$, respectively. If we assume
further that $\bm \Sigma_k = \bm \Sigma$, $k = 1, \ldots, K$, then the pooled
sample covariance matrix $\widehat{\bm \Sigma}$ is substituted for $\widehat{
  \bm \Sigma}_k$ in \eqref{eq:qda}, where
\begin{align}
	\widehat{\bm \Sigma} = N^{-1} \sum_{k=1}^K n_k \widehat{\bm \Sigma}_k \label{eq:pooled-cov}
\end{align}
is the MLE for $\bm \Sigma$. Here, \eqref{eq:qda} reduces to the sample
\emph{LDA} classifier. We omit the log-determinant because it is constant across
the $K$ classes.

@@

Several regularization methods, such as the methods considered by
\cite{Xu:2009fl}, \cite{Guo:2007te}, and \cite{Mkhadri:1995jp} have been
proposed in the literature to adjust the eigenvalues of $\widehat{\bm \Sigma}_k$
so that \eqref{eq:qda} is calculable and has reduced variability in the standard
bias-variance tradeoff. A common form of the covariance-matrix regularization
applies a shrinkage factor $\gamma > 0$, so that
\begin{align}
	\widehat{\bm \Sigma}_k(\gamma) = \widehat{\bm \Sigma}_k + \gamma \bm I_p, \label{eq:ridge-estimator}
\end{align}
similar to a method employed in ridge regression \citep{Hoerl:1970cd}. Equation
\eqref{eq:ridge-estimator} effectively \emph{shrinks} the sample covariance
matrix $\widehat{\bm \Sigma}_k$ toward $\bm I_p$, thereby increasing the
eigenvalues of $\widehat{\bm \Sigma}_k$ by $\gamma$. Specifically, the zero
eigenvalues are replaced with $\gamma$, so that \eqref{eq:ridge-estimator} is
positive definite. For additional covariance-matrix regularization methods, see
\cite{Ramey:2013ji}, \cite{Xu:2009fl}, and \cite{Ye:2009gd}.

@@

Here, we define the \emph{HDRDA} classifier by first formulating the
covariance-matrix estimator $\widehat{\bm \Sigma}_k(\lambda)$ and demonstrating
its clear interpretation as a linear combination of the crossproducts of the
training observations centered by their respective class sample means. We define
the convex combination
\begin{align}
  \widehat{\bm \Sigma}_k(\lambda) \coloneqq (1 - \lambda) \widehat{\bm \Sigma}_k + \lambda
  \widehat{\bm \Sigma}, \quad k = 1, \ldots, K,\label{eq:sig-lambda-alternative}
\end{align}
where $\lambda \in [0, 1]$ is the \emph{pooling} parameter. By rewriting
\eqref{eq:sig-lambda-alternative} in terms of the observations $ \bm x_i$ $(i =
1, \ldots, N)$, each centered by its class sample mean, we attain a clear
interpretation of $\widehat{\bm \Sigma}_k(\lambda)$. That is,
\begin{align}
	\widehat{\bm \Sigma}_k(\lambda)
  &= \left( 1 - \lambda + \frac{\lambda n_k}{N} \right) \widehat{\bm \Sigma}_k +  \frac{\lambda}{N} \sum_{\substack{k' = 1\\k' \ne k}}^K n_{k'} \widehat{\bm \Sigma}_{k'} \nonumber \\
	&= \left( \frac{1 - \lambda}{n_k} + \frac{\lambda}{N} \right)\sum_{i=1}^N
  I(y_i = k) \bm x_i \bm x_i^{T} +  \frac{\lambda}{N} \sum_{i=1}^N I(y_i \ne k)
  \bm x_i \bm x_i^{T} \nonumber \\
	&= \sum_{i=1}^N c_{ik}(\lambda) \bm x_i \bm x_i^{T},\label{eq:sig-lambda-alternative2}
\end{align}
where $c_{ik}(\lambda) = \lambda N^{-1} + (1 - \lambda)n_k^{-1}I(y_i = k)$.
From \eqref{eq:sig-lambda-alternative2}, $\lambda$ weights the contribution of
each of the $N$ observations in estimating $\bm \Sigma_k$ from all $K$ classes
rather than $n_k$ observations from a single class. As a result, we can
interpret \eqref{eq:sig-lambda-alternative2} as a covariance-matrix estimator
that borrows from $\widehat{\bm \Sigma}$ in \eqref{eq:pooled-cov} to estimate $
\bm \Sigma_k$.

@@

As we have discussed above, several eigenvalue adjustment methods have been
proposed to increase eigenvalues (approximately) equal to 0. To further improve
the estimation of $\bm \Sigma_k$ and to stabilize the estimator's inverse, we
define the eigenvalue adjustment of \eqref{eq:sig-lambda-alternative} as
\begin{align}
	\tilde{\bm \Sigma}_k \coloneqq \alpha_k \widehat{\bm \Sigma}_k(\lambda) + \gamma \bm I_p,\label{eq:hdrda-cov}
\end{align}
where $\alpha_k \ge 0$ and $\gamma \ge 0$ is an eigenvalue-shrinkage
constant. Thus, the \emph{pooling} parameter $\lambda$ controls the amount
borrowed from $\widehat{\bm \Sigma}$ to estimate $\bm \Sigma_k$, and the
\emph{shrinkage} parameter $\gamma$ determines the degree of eigenvalue
shrinkage. The choice of $\alpha_k$ allows for a flexible formulation of
covariance-matrix estimators. For instance, if $\alpha_k = 1$, $k = 1, \ldots,
K$, then \eqref{eq:hdrda-cov} resembles \eqref{eq:ridge-estimator}. Similarly,
if $\alpha_k = 1 - \gamma$, then \eqref{eq:hdrda-cov} has a form similar to the
\emph{RDA} classifier from \cite{Friedman:1989tm}. Substituting
\eqref{eq:hdrda-cov} into \eqref{eq:qda}, we define the \emph{HDRDA} classifier
\begin{align}
	D_{HDRDA}(\bm x) = \argmin_{k}  (\bm x - \xbar_k)^{T}\tilde{\bm \Sigma}_k^{+}(\bm x - \xbar_k)  + \log |\tilde{\bm \Sigma}_k|. \label{eq:hdrda}
\end{align}
For $\gamma > 0$, $\tilde{\bm \Sigma}_k$ is nonsingular such that $\tilde{\bm
  \Sigma}_k^{-1}$ can be substituted for $\tilde{\bm \Sigma}_k^{+}$ in
\eqref{eq:hdrda}. If $\gamma = 0$, we explicitly set $|\tilde{\bm \Sigma}_k|$
equal to the product of the positive eigenvalues of $\tilde{\bm
  \Sigma}_k$. Following \cite{Friedman:1989tm}, we select $\lambda$ and $\gamma$
from a grid of candidate models via cross-validation. We provide an
implementation of \eqref{eq:hdrda} in the {\tt hdrda} function in the {\tt
  sparsediscrim} R package, available on CRAN.

@@

Next, we establish properties of the covariance-matrix estimator and the
decision rule employed in the \emph{HDRDA} classifier. By doing so, we
demonstrate that \eqref{eq:hdrda} lends itself to a more efficient
calculation. We decompose \eqref{eq:hdrda} into a sum of two components, where
the first summand consists of matrix operations applied to low-dimensional
matrices and the second summand corresponds to the null space of $\widehat{\bm
  \Sigma}$ in \eqref{eq:pooled-cov}. We show that the matrix operations
performed on the null space of $\widehat{\bm \Sigma}$ yield constant quadratic
forms across all classes and can be omitted. For $p \gg N$, the constant
component involves determinants and inverses of high-dimensional matrices, and
by ignoring these calculations, a substantial reduction in computational costs
is achieved. Furthermore, a byproduct is that adjustments to the associated
eigenvalues have no effect on \eqref{eq:hdrda}. Lastly, we utilize the singular
value decomposition to efficiently calculate the eigenvalue decomposition of
$\widehat{\bm \Sigma}$, further reducing the computational costs of the
\emph{HDRDA} classifier.

@@

\subsubsection{Classification Results}

Similar to \cite{Witten:2011kc}, we report the average test error rates obtained
over 250 random training-test partitions in Table \ref{tab:microarray-results}
along with standard deviations of the test error rates in parentheses. The
\emph{HDRDA} and Guo classifiers were superior in classification performance for
the majority of the simulations with the popular Guo classifier providing
superior classification performance for the Burczysnki data set. Both
\emph{HDRDA} methods yielded comparable results.
